---
title: "Data Preparation"
author: "Peter Menzies"
date: '2022-04-27'
output: html_document
---

# Introduction & Setup

The purpose of this notebook is to provide an example of an analysis on only categorical variables in a RHESSys dataset and provide guidance on how to properly select these variables.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
```

# Data Preparation

## Load Data

```{r load_data, message = FALSE}
df <- readRDS(here::here("data", "input", "sageres.RDS")) 

# Fix scen column
df <- df %>% 
  mutate(scen = ifelse(scen %in% (6:10), scen - 5, scen))
```

## Clean and Aggregate Data

First, we specify the column name for water year, which is the level of aggregation used within this workflow. Next, we specify the categorical variables to be analyzed, which are selected as stratumID, clim, and scen.

Categorical variables that were intentionally left out of the analysis were other spatial IDs (i.e., basinID or hillID), topo, aspect, slope, and elevation. The reason for this is that these variables are all extremely similar to stratumID. The observation's strata provides information on the site's static characteristics and topography. Keeping these variables in addition would lead to biased results of the random forest model. Additionally, random forest models perform better with fewer categorical variables.

```{r user_inputs}
wy_column <- "wy"
categorical_columns <- c("stratumID", "clim", "scen")
response_column <- "npp"
```

```{r input_tests}
# Check class types
if (class(wy_column) != "character") {
  stop("The group columns specified above should be written as characters.")
}
if (class(categorical_columns) != "character") {
  stop("The factor columns specified above should be written as characters.")
}
if (class(response_column) != "character") {
  stop("The response variable column specified above should be written as a character.")
}

# Check for factors with many categories
for (column in categorical_columns) {
  num_categories = n_distinct(df[,column])
  if (num_categories > 50) {
    warning(paste(column, "has", num_categories, "categories, should this column be numeric?"))
  }
  rm(column, num_categories)
}
```

Here is where the categorical variables from the `categorical_columns` object (explained above) get converted to factors. This is necessary for the Random Forest procedure to run. 

```{r convert_to_factors}
# Convert categorical variables to factors
df[,categorical_columns] <- lapply(df[,categorical_columns], factor)

```

The steps needed to prepare a dataset with only categorical columns are as follows:

```{r prepare_data}
# Select categorical variables only
df_cats <- df %>%
  rename("response" = response_column) %>% 
  select(response, all_of(categorical_columns), all_of(wy_column))

# Group by chosen columns
df_cats <- df_cats %>%   
  group_by(across(all_of(c(categorical_columns, wy_column)))) %>%  
  summarise_if(is.numeric, mean) %>% 
  ungroup()

# Reorder response variables first and remove any unwanted variables
df_cats <- df_cats %>% 
  select(response, everything())

# Remove wy column
df_cats <- df_cats %>% 
  select(-wy)
```

```{r}
save.image(here::here("data", "input", "sagehen_cats.RData"))
```

