---
title: "Data Preparation"
author: "Alex Clippinger, Shale Hunter, Peter Menzies, Wylie Hampson"
date: '2022-04-27'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(RHESSysIOinR)
```

## Read in Data

The specific `read` function below may need to be changed below depending on the file type of your dataset.

```{r read_data, message = FALSE}
#                                                 ↓ Enter data file name
df_raw <- read.table(here::here("data", "input", " "), header = TRUE) 
```

## Clean and Aggregate Data

First, RHESSys output will need to be aggregated and cleaned based on specific research question. Possible changes include:

1.  Changing the temporal resolution (i.e. Daily to Yearly measurements).

2.  Converting variable units (i.e. Radians to Degrees).

3.  Converting variable class (i.e. numeric/character to factor).

4.  Creating derived variables (i.e. Peak SWE).

This workflow defaults to aggregating by water year---finer temporal resolutions (e.g. daily, weekly) can be used, but model training and tuning times (particularly for gradient boosting) may increase dramatically depending on compute resources. The code chunk below identifies the variables in our data set that will be converted to factors and used for grouping, and the desired response variable. Change the inputs below to apply to your dataset in whatever way necessary (you might want to include variables that differentiate the conditions used for a single RHESSys simulation---examples might be treatment type, location, climate model scenario). 

The `factor_vars` object should include the predictor variables that will be converted to factors, which allows the random forest workflow to function. Any predictor variables that should be factors instead of numeric or character should be converted. (Example: stratumID is numeric, but because there are 6 different strata, it makes more sense for them to be factors.) Factor variables will be used to aggregate the dataset along with water year or other temporal resolution.

The `response_var` object is the response variable that the user wants to predict. There can only be one response variable.


```{r user_inputs}
# Adding water year
df_raw <- add_dates(df_raw) %>% 
  select(-yd)

#                       ↓ enter factor variable names 
factor_vars <- c("wy", " ")

#                ↓ enter response variable name
response_var <- " "
```

```{r input_tests}
# Check class types
if (class(factor_vars) != "character") {
  stop("The factor columns specified above should be written as characters.")
}
if (class(response_var) != "character") {
  stop("The response variable column specified above should be written as a character.")
}

# Check for factors with many categories
for (column in factor_vars) {
  num_categories = n_distinct(df_raw[,column])
  if (num_categories > 50) {
    warning(paste(column, "has", num_categories, "categories, should this column be numeric?"))
  }
  rm(num_categories, column)
}
```

Because temporal aggregation of the data set is critical to the results, we first create a function that allows us to dynamically select whether monthly or seasonal average temperatures are included as predictor variables. Setting `resolution = 'season'` does seasonal average temperatures, and `resolution = 'month'` does monthly average temperatures. By default, this function is set up to do seasonal average temperatures, and the seasons are:

Winter = Dec, Jan, Feb, March
Spring = April, May
Summer = June, July, Aug, Sep
Fall = Oct, Nov

It is possible to reassign different months to different seasons within the function as well. 

```{r aggregate_temp_function}
aggregate_temp <- function(df, 
                           resolution = 'season', 
                           winter = c(12, 1, 2, 3),
                           spring = c(4, 5),
                           summer = c(6, 7, 8, 9),
                           fall = c(10, 11)) {
  
  if (resolution == 'month') {
    months <- c('jan', 'feb', 'mar', 'apr', 'may', 'jun', 'jul', 'aug', 'sep', 'oct', 'nov', 'dec')
    for (i in seq_along(months)) {
      column_name <- paste0(months[i], '_tavg')
      df_temp_agg[[column_name]] <- mean(df_temp_agg$tavg[df_temp_agg$month == i])
    }
  }
  
  if (resolution == 'season') {
    df_temp_agg <- df %>%
      mutate(winter_tavg = mean(tavg[month %in% (winter)]),
             spring_tavg = mean(tavg[month %in% (spring)]),
             summer_tavg = mean(tavg[month %in% (summer)]),
             fall_tavg = mean(tavg[month %in% (fall)])) 
  }
  
  return(df_temp_agg)
  
  warning("If the season arguments are left blank, this function defaults to winter = c(12, 1, 2, 3), spring = c(4, 5), summer=c(6, 7, 8, 9), fall = c(10, 11)")
}
```

Here is where the categorical variables from the `factor_vars` object (explained above) get converted to factors. This is necessary for the Random Forest procedure to run. 

```{r convert_to_factors}
# Convert categorical variables to factors
df_raw[, factor_vars] <- lapply(df_raw[, factor_vars], factor)

```

Next, the data are aggregated and averaged by water year and other factor variables.

```{r prepare_data}
# Rename chosen response variable to "response"
df_raw <- df_raw %>% 
  rename("response" = response_var)

# Remove columns with zero variance
num_vars <- sapply(df_raw, is.numeric)
df_num_vars <- df_raw[, num_vars]
no_variance <- sapply(df_num_vars, function(col) var(col) == 0)
no_variance <- colnames(df_num_vars[, no_variance])
df_raw <- df_raw %>% 
  select(!all_of(no_variance))
rm(num_vars, df_num_vars, no_variance)

# Group by chosen columns
df <- df_raw %>%   
  group_by(across(all_of(factor_cols)))

# Aggregate average temperatures using the aggregate_temp() function defined above.
df <- aggregate_temp(df, 'season')

# Average all numeric values by groups
df <- df %>% 
  summarise_if(is.numeric, mean) %>% 
  ungroup()

# Reorder response variables first and remove dates
df <- df %>% 
  select(response, everything()) %>%
  select(-c(wy, day, month, year))
```


## Split Data (optional)

The code below offers the option to split the current dataframe into multiple by a given variable. Doing so allows changes in variable relationships to be observed across different groups. The case study around which this workflow was designed involved different climate warming scenarios and separate dataframes were used to assess changes across those scenarios. This step is not necessary, however.

If this is not applicable to your dataset, this step can be ignored and will not run.

If you would like to split the dataframe, set the `split_df` object below to `TRUE`, and assign the name of the desired variable to the `split_var` object.


```{r split_data}
split_df <- FALSE

if (split_df == TRUE) {
  
#               ↓ enter variable to split by 
  split_var <- " "
  split_groups <- levels(df[[split_var]])
  
  if (length(split_groups) > 4) {
    stop("It's advised not to create more than 4 different dataframes")
  }
  
  for (i in seq_along(split_groups)) {
    tmp_df <- df %>% 
      filter(split_var == split_groups[i])
    
    assign(tmp_df, paste0("df_", split_groups[i]))
  }
  
  # remove temporary objects
  rm(c(tmp_df, climate_scenarios))
}  
```

## Save Data

Here we save all dataframes created during this process. 

Note: `save.image()` saves all objects currently in the RStudio environment, so if you have objects loaded unrelated to this workflow they will be saved in the file created below as well.

```{r}
save.image(here::here("data", "input", "prepared_data.RData"))
```

