---
title: "Variable Importance Template - Random Forest"
author: "Alex Clippinger, Wylie Hampson, Shale Hunter, Peter Menzies"
date: '2022-06-03'
output: 
  rmarkdown::html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: cerulean
---

# Introduction

[RHESSys](https://github.com/RHESSys/RHESSys/wiki) output provides multiple variables that describe the response of ecosystem biogeochemical cycling and hydrology to climate (and land use) drivers. This R Markdown document describes a workflow to apply machine learning techniques to RHESSys model output. This particular notebook uses random forest---see the other document for gradient boosting. The specific goal is to determine the most important relationships between RHESSys predictor variables and a chosen response variable in a programmatic, efficient, and reproducible manner. This is done by assessing variable importance within random forest models and the metric: permutation importance. 

No machine learning experience is required to make use of this workflow---the details of random forest and permutation importance will be explained throughout the course of the document.

## Data Preparation Step:

If you have not already, use the `data_preparation_template.Rmd` to prepare your data before working through this notebook.


# Setup

## renv:

To help ensure reproducibility, the packages and package versions used to build this workflow have been saved via [`renv`](https://rstudio.github.io/renv/articles/renv.html). If this is your first time using the workflow and you have not yet run `renv::restore()`, please set `renv_restore` in the setup chunk below to `TRUE`. This will install any necessary packages and load the versions that were used to create the notebook. Doing so will not impact your package versions outside of this R Project. It may take several minutes depending on the number of discrepancies.

## Load Packages

```{r setup, include = TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)

# renv setup
librarian::shelf("renv")
#                 ↓ set to TRUE if you have not already run `renv::restore()`
renv_restore <- FALSE

if (renv_restore == TRUE) {
  renv::restore()
}

# Standard packages
library(tidyverse)
library(here)
library(patchwork)
library(psych)
library(kableExtra)
library(zeallot)
library(DT)

# Machine Learning packages
library(caret)
library(spatialRF)
library(randomForest)
library(party)
library(partykit)
library(rfUtilities)
library(randomForestExplainer)
```

# Data

## Load Prepared Data

Here, we load the data files created in the `data_preparation_template.Rmd`.

```{r}
load(here::here("data", "input", "prepared_data.RData"))
```

## Summarize Data

Next, we want to get a summary description of the data set. This step can be used to check if any variables do not have the expected range, magnitude, class, etc.


```{r describe_data}
source(here::here("R", "summarize_data.R"))

summarize_data(df)
```


# Remove Multicollinearity

Highly correlated predictor variables are not as much a concern in machine learning when creating a predictive model. However, for this process of assessing relative predictor variable importance, multicollinear variables have biased importance (Strobl et al. 2008). Therefore, these need to be handled prior to assessing feature importance.

## Method

Below, we use Variance Inflation Factors (VIF) and Pearson Correlation Coefficients to remove variables with high multicollinearity. This is done using `auto_vif()` and `auto_cor()` from the `spatialRF` package---these two functions are used in conjunction to best minimize multicollinearity. Both functions allow the user to define an order of preference for the selection of variables, which will be discussed below. If no preference order is decided, the `auto_vif()` function orders the variables from minimum to maximum VIF, and the `auto_cor()` function orders variables by their column order.

The `auto_vif()` function works by initially starting with the variable with highest preference. Then, it iterates through the preference order list, computing the VIF for each new variable added. If the VIF of the new variable is lower than the threshold it is kept, if not the variable is removed. This process is continued until all VIF values are below the user-input threshold.

Similarly, the `auto_cor()` function works by computing a correlation matrix for all variables. Next, it computes the maximum correlation for each variable. The function then begins with the variable with lowest preference. If that variables maximum correlation coefficient is above the user-input threshold, that variable is removed. This process is continued until all correlation values are below the user-input threshold.

## Identify and Remove Correlated Variables

First, we create data frames containing only the predictor variables to assist with the next steps.

```{r get_predictors}
# Save data frames of predictor variables
df_num_preds <- df %>% 
  select(!response & where(is.numeric))

df_fpreds <- df %>% 
  select(!response & where(is.factor)) %>% 
  colnames()
```

Below, there are two methods for setting preference order: (1) Manually creating an ordered vector of column names, or (2) Allowing a preliminary random forest method to determine preference order based on variable importance. Note that in the preliminary random forest, highly correlated variables will not produce accurate estimates of importance. However, we assume that relative importance is reasonably accurate to support the selection between highly correlated variables. The second method is used by default. 

1. Run this chunk to manually create an ordered vector of column names, starting with those of highest priority / greatest interest. The vector does not need to include the names of all predictors - only those that you would like to keep in the analysis. Set `set_manual_pref_order` to `TRUE` to run.

```{r manual_preference_order}
# OPTIONAL step
# Preference order can be set manually for variables of interest:

set_manual_pref_order <- FALSE

if (set_manual_pref_order == TRUE) {

  # set to variables of interest if using manual preference order
  pref_order <- c("precip", "rz_storage", "trans", "evap")
  
}


```

2. Run this chunk to allow a preliminary random forest to automatically determine preference order.

```{r auto_preference_order, warning=FALSE}
# Find preliminary importance using random forest
prelim_imp <- train(x = df %>% select(!response),
              y = df$response,
              method = 'rf',
              importance = TRUE,
              replace = TRUE,
              trControl = trainControl(method = "none", seed = 4326))

# Set preference order based on variable importance
pref_order <- varImp(prelim_imp$finalModel, scale = FALSE) %>% 
  arrange(-Overall) %>% 
  rownames()
```

Thresholds for VIF and correlation can be set below, with default values of 5 and 0.75, respectively. Increasing the thresholds will reduce the number of variables that get removed, but it will increase the likelihood that collinearity influences the results.

```{r set_thresholds}
vif.threshold = 5
cor.threshold = 0.8
```

```{r remove_multicollinearity, warning = FALSE}
source(here::here("R", "remove_vif.R"))
source(here::here("R", "remove_cor.R"))

# Identifying variables with correlations or VIF above given thresholds
vif_cor <- remove_cor(df_num_preds, cor.threshold = cor.threshold, pref_order) %>% 
  remove_vif(vif.threshold = vif.threshold, pref_order)

select_variables <- vif_cor$selected.variables

# Remove numeric variables with multicollinearity
df_reduced <- df %>% 
  select(c(response, all_of(df_fpreds), all_of(select_variables)))
```

## Summary of Removed Variables

The next step is intended to elucidate the previous multicollinearity reduction process via figures showing which variables were selected and removed, and why. This information can be used to determine whether the auto-generated preference order based on preliminary importance performed satisfactorily, or whether the preference order should be set manually. If, for example, a variable of particular interest was removed, the `manual_preference_order` code chunk above can be used to create a preference order containing the desired variable.


```{r summarize_cor_function, class.source='fold-hide'}
source(here::here("R", "summarize_var_removal.R"))
source(here::here("R", "plot_removed_cor.R"))
```

The following functions output summary tables and plots showing variable selection status and the values of VIF, correlation, and importance which led to those selections. 

### Summary Tables {.tabset}

#### Variable Removal Summary Table

```{r vif_summary_table_0, warning=FALSE}
summarize_var_removal(df, select_variables, prelim_imp, table = TRUE)
```

#### Cor Summary Plot

```{r}
plot_removed_cor(select_variables, df_num_preds)
```

# Feature Importance with Random Forest

With multicollinearity reduced, the importance of predictor variables can be estimated using random forest and partial permutation importance. In this framework, we use random forest because it has been shown to be an effective tool in assessing feature importance in numerous applications including ecological analysis (Cutler et al. 2007; Prasad et al. 2006). Additionally, random forest requires relatively less hyper-parameter tuning than other common techniques such as gradient boosting, which is explored in separate workflows. 

### Hyper-Parameter Tuning

An essential part of building a machine learning model is tuning hyper-parameters. This entails altering parameters which the user has control over in order to attain an optimal model accuracy. The following tuning process has been automated and only involves `mtry`---the primary parameter that requires adjustment in a random forest.

The `mtry` parameter determines the number of variables randomly sampled as candidates at each split. The default value for regression is p/3, where p is the number of predictor variables.

The function below creates a vector of accuracy results of random forests using different values for `mtry`.

```{r tuning_function}
df_reduced <- as.data.frame(df_reduced)
```

Next, we run the function for both climate scenario data sets and visualize the results. Model error here is assessed using out-of-bag (OOB) observations in the data, which refers to observations that were not sampled during bootstrapping for the particular random forest being assessed. In other words OOB data were not used to train the model and can be used to test accuracy without the risk of bias incurred from model over-fitting. The best `mtry value` in the plots below will be represented by the lowest `OOB Error`.

```{r tune_rf, warning = FALSE}
source(here::here("R", "tune_rf_model.R"))

set.seed(4326)

mtry <- tune_rf_model(df_reduced)
bestmtry <- match(min(mtry), mtry)
ggplot(data = as.data.frame(mtry), aes(x = 1:length(mtry), y = mtry)) +
  geom_vline(xintercept = bestmtry, linetype = "dashed") +
  geom_point() +
  geom_line() +
  theme_light() +
  labs(x = "mtry value",
       y = "OOB Error",
       title = paste0("The best mtry value for ", deparse(substitute(df_reduced)), " is ", bestmtry))
```

The best `mtry` value is used random forest model below.

### Random Forest Models

```{r get_random_forests}
rf <- train(x = df_reduced %>% select(!response),
                y = df_reduced$response,
                method = 'rf',
                .mtry = bestmtry,
                ntree = 500,
                importance = TRUE,
                replace = TRUE,
                keep.forest = TRUE,  # necessary for permimp
                keep.inbag = TRUE,  # necessary for permimp
                trControl = trainControl(method = "none", seed = 4326)) # to use 10-fold cross validation, method = "cv", number = 10
```

Assessing feature importance is a complex task with many possible approaches. Tree based models like random forest offer convenient "split-improvement" measures, like mean increase in purity and minimum depth, which are intrinsically derived during model construction. However, these have been shown to be biased towards variables with many categories or large value ranges (Strobl et al. 2007). Despite some of their shortcomings, these importance measures can provide insights are further explored at the end of this document.

As the primary measure of importance we instead use partial permutation importance via the `varImp` function. Permutation importance has been tested and determined to be the most unbiased importance metric for data with a mix of categorical and continuous variables with a variety of classes and ranges, respectively. This is calculated in the following steps...

1. Assess prediction accuracy (mean squared error) of the model on the out-of-bag data.

2. Permute the values of a given variable.

3. Feed the dataset with the permuted values into the random forest and reassess model accuracy.

4. Importance of the permuted variable is deemed to be the mean loss in accuracy when its values were permuted. 

```{r variable_importance}
imp <- varImp(rf$finalModel, scale = FALSE) %>% 
  rownames_to_column("Variable") %>% 
  mutate(Rank = rank(-Overall))
```


# Model Evaluation

The following section provides visualizations and statistics to evaluate the random forest performance.

First, simply calling the random forest shows a summary of the results, including percent variance explained.

```{r}
rf_fit <- round(tail(rf$finalModel$rsq, 1)*100, 2)

rf$finalModel
```

The above model explained `r rf0_fit`% of variance in the response variable.


# Visualize Results

Now, we can visualize the results of the random forest feature selection.

Variable importance plots display relative magnitude of importance between variables.

```{r plot_imp}
source(here::here("R", "plot_imp.R"))

plot <- plot_imp(imp)
plot
```


# Further Exploration

## Shiny App

The capstone team created an application with a suite of interactive visualization tools using the `Shiny` package. It can be used to further explore the original data set and the variable importance results from the workflow. The app can be launched below by setting `run_app` below to `TRUE`.

First, the following objects are saved for use in the app.

```{r save_model_output}
save(rf, imp, file = here::here("data", "output", "model_output.RData"))
```

```{r}
run_app <- FALSE

if (run_app == TRUE) {
  shiny::runApp(here::here("shiny"))
}
```


## Additional Measures of Importance

While permutation importance was chosen as the primary measure of importance in this workflow, there are other measures that random forests allow for which may be of utility. It's important, however, that the inherent biases previously mentioned (Strobl et al. 2007) are understood. These additional measures will not be described further here, but the function below, `randomForestExplainer::explain_forest()`, is an excellent tool which creates an html document visualizing and comparing a number of them.

To run the explainer, just set the `run_explainer` object below to `TRUE`. It will likely take several minutes to run. The document is set to output to the "docs" folder in the project's root directory, and will be best viewed in your browser.

*Note: the measure of "mse_increase" found in the output document is synonymous with "permutation importance" used in the workflow.*

```{r}
# Set to false by default due to long run time
run_explainer <- FALSE

if (run_explainer == TRUE) {
  randomForestExplainer::explain_forest(rf0, path = here::here("docs", "forest_explained.html"))
}
```



# References

Cutler, D. Richard, Thomas C. Edwards, Karen H. Beard, Adele Cutler, Kyle T. Hess, Jacob Gibson, and Joshua J. Lawler. 2007. "Random Forests for Classification in Ecology." *Ecology* 88 (11): 2783--92. <https://doi.org/10.1890/07-0539.1>.

Debeer, Dries, and Carolin Strobl. 2020. "Conditional Permutation Importance Revisited." *BMC Bioinformatics* 21 (1): 307. <https://doi.org/10.1186/s12859-020-03622-2>.

Prasad, Anantha M., Louis R. Iverson, and Andy Liaw. 2006. "Newer Classification and Regression Tree Techniques: Bagging and Random Forests for Ecological Prediction." *Ecosystems* 9 (2): 181--99. <https://doi.org/10.1007/s10021-005-0054-1>.

Strobl, Carolin, Anne-Laure Boulesteix, Achim Zeileis, and Torsten Hothorn. 2007. "Bias in Random Forest Variable Importance Measures: Illustrations, Sources and a Solution." *BMC Bioinformatics* 8 (1): 25. <https://doi.org/10.1186/1471-2105-8-25>.

Strobl, Carolin, Anne-Laure Boulesteix, Thomas Kneib, Thomas Augustin, and Achim Zeileis. 2008. "Conditional Variable Importance for Random Forests." *BMC Bioinformatics* 9 (1): 307. <https://doi.org/10.1186/1471-2105-9-307>.
